The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-08 dictionary.tex
Title

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[Lost in Translation: The function--sheaf dictionary]{Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences, University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{sheaf--function dictionary, geometrization, quasicharacter sheaves, commutative group schemes}

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes, showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

\subsection*{Main result}

@@

So, what is a quasicharacter sheaf?
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.
The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section,
and the kernel of the Trace of Frobenius is classified by $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$.
In particular, in contrast to the case of connected commutative algebraic groups, the Trace of Frobenius may not injective.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
The canonical section of the Trace of Frobenius allows one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$.
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

\subsection*{Illustrative examples}

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We also thank Takashi Suzuki, who

@@

The category $\QC(G)$ of quasicharacter sheaves on $G$ is a rigid monoidal category
\cite{etingof:09a}*{\S1.10} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf hom functor
$\sheafHom(\ - \ ,\EE)$.
The category of quasicharacter sheaves on $G$ is not abelian, so $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}*{0.1}.
This rigid monoidal category structure for $\QC(G)$ gives the set $\QCiso{G}$
of isomorphism classes in $\QC(G)$ the structure of a group.

@@

The proof of Proposition~\ref{prop:functorialG} will be given at the end of Section~\ref{sec:pullback}.

@@

@@

\begin{proposition} \label{prop:etale-iso}
	If $G$ is an \'etale commutative group scheme over $\Fq$
	and if $G(\bFq)$ is finitely generated,
  	then
  	\[
  	\QCiso{G} \iso \Hh^2(G(\bFq)\rtimes \Weil{}, \EEx).
  	\]
  	Moreover, this isomorphism is functorial
  	in the sense of Proposition~\ref{prop:functorialG}.
\end{proposition}

\begin{proof}
By Lemma~\ref{lem:dual-inv}, $G(\Fq)^* \iso \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))$. Composing this isomorphism with the Trace of Frobenius gives
\[
T_{G}: \QCiso{G}\to  \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx)),
\]
which is surjective, Lemma~\ref{lem:etale-trace}.
Combining this with
\[
S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))
\]
gives
\[
\QCiso{G} \mathop{\longrightarrow}\limits^{T_{G}\oplus S_{G}} \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))\oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
The Hochschild-Serre spectral sequence (for finitely generated abelian groups!) provides
\[
\Hh^2(G(\bFq)\rtimes \Weil{},\EEx)  \iso \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx)) \oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
Composing these last two maps defines
\begin{equation}
\QCiso{G}\to \Hh^2(G(\bFq)\rtimes \Weil{},\EEx).
\end{equation}
To see that this homomorphism is an isomorphism,\todo{Clifton, you are here.}
\end{proof}

@@

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme,
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
%0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$
in Sections~\ref{sec:connected} and \ref{sec:etale}.
In this section we show that \eqref{eq:pi2} is exact.

@@

\begin{proposition}\label{prop:middleexact}
 The sequence
 \[
  \begin{tikzcd}
%  0 \rar &
 \QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
  \end{tikzcd}
 \]
 is exact.
\end{proposition}

\begin{proof}
We already know the sequence is exact at $\QCiso{G^0}$.
Here we show exactness at $\QCiso{G}$.
First note that $\iota_0^* \circ \pi_0^*$ is trivial by Proposition~\ref{prop:pullback}.
So it suffices to show that if $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$
with $\qcs{L}\vert_{G^0} = (\EE)_{G^0}$ then $\qcs{L}$ is in the essential image of $\pi_0^*$.

@@

\section{The function--sheaf dictionary for smooth, commutative group schemes}
\label{sec:snake}

>>>>

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[The function--sheaf dictionary]{The function--sheaf dictionary for smooth commutative group schemes over finite fields}

\title[Lost in Translation: The function--sheaf dictionary]{Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields}

\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences, University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{sheaf--function dictionary, geometrization, quasicharacter sheaves, commutative group schemes}

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes}%showkeys}
\usepackage{amsmath, amsthm, amssymb, todonotes, showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

Let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
\subsection*{Main result}

@@

So, what is a quasicharacter sheaf?
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.
The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section,
and the kernel of the Trace of Frobenius is classified by the classes in $\Hh^2(\pi_0({\bar G}),\EEx)$ that are fixed by Frobenius.
In contrast to the case of connected commutative algebraic groups, the Trace of Frobenius is not injective.
and the kernel of the Trace of Frobenius is classified by $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$.
In particular, in contrast to the case of connected commutative algebraic groups, the Trace of Frobenius may not injective.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
The canonical section of the Trace of Frobenius allows one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$.
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

Three examples will illustrate some of the issues that arise when extending the function--sheaf dictionary to smooth commutative group schemes.
\subsection*{Illustrative examples}

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We also thank Takashi Suzuki, who 
We also thank Takashi Suzuki, who

@@

The category $\QC(G)$ of quasicharacter sheaves on $G$ is a rigid monoidal category
\cite{etingof:09a}*{\S1.10} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf hom functor
$\sheafHom(\ - \ ,\EE)$.
The category of quasicharacter sheaves on $G$ is not abelian, so $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}*{0.1}.
This rigid monoidal category structure for $\QC(G)$ gives the set $\QCiso{G}$
of isomorphism classes in $\QC(G)$ the structure of a group.
We now begin developing the tools to relate $\QCiso{G}$ to $G(\Fq)^*$.

@@

The proof of Proposition~\ref{prop:functorialG} will be given at the end of the next section.
The proof of Proposition~\ref{prop:functorialG} will be given at the end of Section~\ref{sec:pullback}.

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined below.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $g_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that the class that $g_\mathcal{L} = g_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
\QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $g_\qcs{L}$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
f(x+y) = f(x) f(y) \frac{g(\Frob{G}(x),\Frob{G}(y))}{g(x,y)}
\end{equation}
for all $x,y \in G(\bFq)$. Thus, the class $\,^{\Frob{}}g_\qcs{L}$ of $\,^{\Frob{}}g$ coincides
with the class $g_\qcs{L}$ of $g$. This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.

If $G$ is a connected and of finite type, 
then $H^2(G(\bFq),\EEx)$ is trivial \todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible; 
 by contrast, if $G$ is etale then $H^2(G(\bFq),\EEx)$ may be non-trivial and $G$ may admit invisible quasicharacters sheaves.
We will return to the latter case in Section~\ref{sec:etale}.
In the meantime, we end this section with a simple property of $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.

\begin{lemma}
The function \[S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))\] is a group homomorphism.
\end{lemma}

\begin{proof}
We must show:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.\todo{Under construction.}
\end{proof}

@@

\begin{proposition} \label{prop:etale-iso}
	If $G$ is an \'etale commutative group scheme over $\Fq$ 
	If $G$ is an \'etale commutative group scheme over $\Fq$
	and if $G(\bFq)$ is finitely generated,
  	then
  	\[
  	\QCiso{G} \iso \Hh^2(G(\bFq)\rtimes \Weil{}, \EEx).
  	\]
  	Moreover, this isomorphism is functorial
  	in the sense of Proposition~\ref{prop:functorialG}.
\end{proposition}

\begin{proof}
By Lemma~\ref{lem:dual-inv}, $G(\Fq)^* \iso \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))$. Composing this isomorphism with the Trace of Frobenius gives 
By Lemma~\ref{lem:dual-inv}, $G(\Fq)^* \iso \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))$. Composing this isomorphism with the Trace of Frobenius gives
\[
T_{G}: \QCiso{G}\to  \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx)),
\]
which is surjective, Lemma~\ref{lem:etale-trace}. 
Combining this with 
which is surjective, Lemma~\ref{lem:etale-trace}.
Combining this with
\[
S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))
\]
gives
\[
\QCiso{G} \mathop{\longrightarrow}\limits^{T_{G}\oplus S_{G}} \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))\oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
The Hochschild-Serre spectral sequence (for finitely generated abelian groups!) provides  
The Hochschild-Serre spectral sequence (for finitely generated abelian groups!) provides
\[
\Hh^2(G(\bFq)\rtimes \Weil{},\EEx)  \iso \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx)) \oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
Composing these last two maps defines
\begin{equation}
\QCiso{G}\to \Hh^2(G(\bFq)\rtimes \Weil{},\EEx).
\end{equation}
To see that this homomorphism is an isomorphism,\todo{Clifton, you are here.}
\end{proof}

@@

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme,
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
%0 \rar & 
%0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$
in Sections~\ref{sec:connected} and \ref{sec:etale}.
In this section we show that \eqref{eq:pi2} is exact.

@@

\begin{proposition}\label{prop:middleexact}
 The sequence
 \[
  \begin{tikzcd}
%  0 \rar & 
%  0 \rar &
 \QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
  \end{tikzcd}
 \]
 is exact.
\end{proposition}

\begin{proof}
We already know the sequence is exact at $\QCiso{G^0}$. 
We already know the sequence is exact at $\QCiso{G^0}$.
Here we show exactness at $\QCiso{G}$.
First note that $\iota_0^* \circ \pi_0^*$ is trivial by Proposition~\ref{prop:pullback}.
So it suffices to show that if $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$
with $\qcs{L}\vert_{G^0} = (\EE)_{G^0}$ then $\qcs{L}$ is in the essential image of $\pi_0^*$.

@@

\section{The function--sheaf dictionary for disconnected group schemes}
\section{The function--sheaf dictionary for smooth, commutative group schemes}
\label{sec:snake}
